---
output: html_document
---
Exploratory Data Analysis of Red Wine dataset by Kriti Singh
========================================================
Citation :
  This dataset is public available for research. The details are described in [Cortez et al., 2009]. 

  P. Cortez, A. Cerdeira, F. Almeida, T. Matos and J. Reis. 
  Modeling wine preferences by data mining from physicochemical properties.
  In Decision Support Systems, Elsevier, 47(4):547-553. ISSN: 0167-9236.

  Available at: [@Elsevier] http://dx.doi.org/10.1016/j.dss.2009.05.016
                [Pre-press (pdf)] http://www3.dsi.uminho.pt/pcortez/winequality09.pdf
                [bib] http://www3.dsi.uminho.pt/pcortez/dss09.bib
                
                
```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages 
# in your analysis in this code chunk.


install.packages("ggplot2", repos="http://cran.rstudio.com/", 
                 dependencies = T) 
install.packages("knitr", repos="http://cran.rstudio.com/",
                 dependencies = T)
install.packages("dplyr", repos="http://cran.rstudio.com/",
                 dependencies = T)


library(ggplot2)
library(knitr)
library(dplyr)
```

```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE)

```
```{r echo=FALSE, message=FALSE, warning=FALSE, Load_the_Data}
# Load the Data
setwd("C:/Users/krs437/Documents/RStudio/Files")
mydata=read.csv("wineQualityReds.csv", sep=',')
  
```
 
 The dataset is "tidy". Therefore, we don't need data wrangling here. 
 
# Univariate Plots Section
```{r echo=FALSE, Structure_of_Data}
str(mydata)
summary(mydata)
  
```

# Univariate Analysis

### Structure of the dataset
Our data frame consists of 1599 observations with 13 variables. Out of the 13 variables, 12 are input variables, while only one, that is, quality is output varibale. 


### main feature of interest in the dataset
The main feature of interest in the dataset is the quality of the red wine. 

###Other features in the dataset that support our investigation
There is only one dependent variable, quality of wine. Therefore, we will try to see how different factors (input variables) affect the quality of wine

```{r echo=FALSE,  message=FALSE, warning=FALSE, Univariate_Plots}
qplot(factor(mydata$quality), geom="bar")
  
```

Plot is a uniform distribution. Therefore, no transformation is necessary for axis of quality. The plot shows that most number of wines lie in the average range, that is, quality score 5-6.
Let's summarise the quality to look at the maximum and minimum value of quality.

```{r echo=FALSE}
summary(mydata$quality)

```

The minimum is 3 and maximum is 8. 

```{r echo=FALSE, message=FALSE, warning=FALSE, Histogram_pH}
ggplot(aes(x=pH), data=mydata)+ 
  geom_histogram()
  

```

For most wines the pH lies bewtween 3.0 and 3.5 with a spike in exactly the middle.

```{r echo=FALSE,  message=FALSE, warning=FALSE,Histogram_volatile.acidity}
ggplot(aes(x=volatile.acidity), data=mydata)+ 
  geom_histogram()
  

```

Let's adjust the binwidth

```{r echo=FALSE, message=FALSE, warning=FALSE, Histogram_volatile.acidity_adjusted_binwidth }
ggplot(aes(x=volatile.acidity), data=mydata)+ 
  geom_histogram(binwidth=0.01)

summary(mydata$volatile.acidity)
    

```


The volatile.acidity seems to be a factor that influences the quality of wine. While most wines lie within the interquartile range. There are many that lie outside. It would be ineresting to see how volatile acidity affects the quaity of wine.

```{r echo=FALSE,  message=FALSE, warning=FALSE, Histogram_fixed.acidity}
ggplot(aes(x=fixed.acidity), data=mydata)+ 
  geom_histogram()
  
summary(mydata$fixed.acidity)

```

We see that fixed acidity ranges between 4.60 and 15.90 with a small IQR (7.10, 9.20). This implies that the fixed acidity doesnt vary much

```{r echo=FALSE,  message=FALSE, warning=FALSE, Histogram_citric.acid}
ggplot(aes(x=citric.acid), data=mydata)+ 
  geom_histogram()
  
summary(mydata$citric.acid)

```
The max value seems to be an outlier, therefore, we should limit our plot.

```{r echo=FALSE,  message=FALSE, warning=FALSE, Histogram_citric.acid_revised}
ggplot(aes(x=citric.acid), data=mydata)+ 
  geom_histogram(binwidth = 0.1)+
  scale_x_continuous(lim=c(0,0.8))
```

It is clear from the histogram that citic acid lies between 0.1 and 0.42 that is the IQR

```{r echo=FALSE,  message=FALSE, warning=FALSE, Histogram_residual_sugar}
ggplot(aes(x=residual.sugar), data=mydata)+ 
  geom_histogram()

summary(mydata$residual.sugar)

```

Most of the values lie in the IQR. However, there is some data that lies on the far end side. 

```{r echo=FALSE,  message=FALSE, warning=FALSE, Histogram_chlorides}
ggplot(aes(x=chlorides), data=mydata)+ 
  geom_histogram()

summary(mydata$chlorides)

```

There are very few values that lie outside IQR. Thereofere, this may not be a influential factor.

```{r echo=FALSE,  message=FALSE, warning=FALSE, Histogram_free_SO2}
ggplot(aes(x=free.sulfur.dioxide), data=mydata)+ 
  geom_histogram()

summary(mydata$free.sulfur.dioxide)

```

```{r echo=FALSE,  message=FALSE, warning=FALSE, Histogram_total.sulfur.dioxide}
ggplot(aes(x=total.sulfur.dioxide), data=mydata)+ 
  geom_histogram()

summary(mydata$total.sulfur.dioxide)

```

Both free and total SO2 are widely distributed. Thereofre, this may be imprtant factor.

```{r echo=FALSE,  message=FALSE, warning=FALSE, Histogram_density}
ggplot(aes(x=density), data=mydata)+ 
  geom_histogram()

summary(mydata$density)

```

density is uniformly distributed with few outliers.

```{r echo=FALSE,  message=FALSE, warning=FALSE, Histogram_sulphates}
ggplot(aes(x=sulphates), data=mydata)+ 
  geom_histogram()

summary(mydata$sulphates)

```

```{r echo=FALSE,  message=FALSE, warning=FALSE, Histogram_alcohol}
ggplot(aes(x=alcohol), data=mydata)+ 
  geom_histogram()

summary(mydata$alcohol)

```

Alcohol has a wide ditribution. Therefore, this seems to be an important factor for our analysis. 


Let's move on and explore how different independent variables are correlated with the dependent variable.

```{r echo=FALSE,  message=FALSE, warning=FALSE,Correlation_quality_with_all_variables}
##the moveme function makes quality the first column of the data. This makes splitting the data for correlation calculation easier

moveme <- function (invec, movecommand) {
  movecommand <- lapply(strsplit(strsplit(movecommand, ";")[[1]], 
                                 ",|\\s+"), function(x) x[x != ""])
  movelist <- lapply(movecommand, function(x) {
    Where <- x[which(x %in% c("before", "after", "first", 
                              "last")):length(x)]
    ToMove <- setdiff(x, Where)
    list(ToMove, Where)
  })
  myVec <- invec
  for (i in seq_along(movelist)) {
    temp <- setdiff(myVec, movelist[[i]][[1]])
    A <- movelist[[i]][[2]][1]
    if (A %in% c("before", "after")) {
      ba <- movelist[[i]][[2]][2]
      if (A == "before") {
        after <- match(ba, temp) - 1
      }
      else if (A == "after") {
        after <- match(ba, temp)
      }
    }
    else if (A == "first") {
      after <- 0
    }
    else if (A == "last") {
      after <- length(myVec)
    }
    myVec <- append(temp, values = movelist[[i]][[1]], after = after)
  }
  myVec
}
mydata_new<-mydata[moveme(names(mydata), "quality first")]
COR <- cor(as.matrix(mydata_new[,1]), as.matrix(mydata_new[,-1]))
COR



```


From the above correlation table, we see that "fixed.acidity", "citic.acid", "residual.sugar", "sulphates" and "alcohol" are positviely correlated to the quality of wine. This means that as these factors move upward, the quality of wine also increases.
On the other hand, "volatile,acidity", "chlorides", "free sulfur dioxide", "total sulfur dioxide", "density"" and "pH" are negatively correlated to quality. We will explore this further. 

In addition to that, the correlation table gives us an idea of how each variable is correlated to quality. And, therefore, I chose to not use the ggpair function to save time. 

###create any new variables from existing variables in the dataset
We can group "free.sulfur.dioxide" and "total.sulfur.dioxide" together since they both measure the effect of free sulfur dioxide gas on the quality of wine. We create a new variable sulfur_dioxide


```{r echo=FALSE,  message=FALSE, warning=FALSE}
mydata$sulfur_dioxide<-mydata$free.sulfur.dioxide + mydata$total.sulfur.dioxide
cor.test(mydata$quality, mydata$sulfur_dioxide)
```

sulfur_dioxide is overall negatively correlated to quality of wine

Now, we are ready to plot 2 variables.

# Bivariate Plots Section

```{r echo=FALSE,  message=FALSE, Include="FALSE", warning=FALSE,quality_vs_alcohol}
ggplot(aes(x=quality, y=alcohol), data=mydata)+ 
  geom_point(alpha=1/5, position = "jitter")+
  geom_smooth(method="lm")
  scale_x_discrete(lim=c(3,8), breaks=seq(3,8,1))
  

```

The scatter plot between quality of wine and alcohol content tells us something about their correlation. The high quality alcohol seems to have more alcohol content. Moreover, the correlation coefficient is 0.48 which is strongest of all correlations with quality. The linear fit also shows a positive trend in the plot. 


Let's try boxplot


```{r echo=FALSE, message=FALSE, Include="FALSE", warning=FALSE, boxplot_quality_vs_alcohol}
plot<-ggplot(aes(x = factor(quality), y = alcohol), data = mydata) +
  geom_jitter(alpha=0.3)+
  geom_boxplot(alpha=0.5,color="blue",
               outlier.color=NA) +
  stat_summary(fun.y="mean", geom="point", color="red",
               shape=10, size=4 )
  
  
  
plot_revised<-plot+scale_colour_brewer(palette = "Sequential") 
plot_revised2<-plot_revised+theme_dark()
plot_revised2
cor.test(mydata$quality, mydata$alcohol)

```


The boxplot gives us a clearer picture that for higher quality wine, the alcohol content is higher. The correlation is 0.476.
For better visualization, let's plot the median. This can be done by first grouping the dataset according to quality:-

```{r echo=FALSE, message=FALSE, Include="FALSE", warning=FALSE, group_by_quality}
wine_by_quality <- mydata %>%
      group_by(quality) %>%
      summarise(mean_alcohol = mean(alcohol),
              median_alcohol = median(alcohol),
              n = n()) %>%
      ungroup() %>%
      arrange(quality)
  

  
ggplot(aes(x = quality, y = median_alcohol), 
       data = wine_by_quality) + 
      geom_bar(stat = 'identity') + 
      scale_x_continuous(breaks = seq(3, 8, 1))
  


```

We can repeat the above steps to understand the effect of important factors (based on higher correlation number). 

```{r echo=FALSE,  message=FALSE, warning=FALSE,quality_vs_volatile.acidity_bar}
plot<-ggplot(aes(x = factor(quality), y = volatile.acidity), data = mydata) +
  geom_jitter(alpha=0.3)+
  geom_boxplot(alpha=0.5,color="blue",
               outlier.color=NA) +
  stat_summary(fun.y="mean", geom="point", color="red",
               shape=10, size=4 )
  facet_wrap(~quality, nrow=1) 
  
plot_revised<-plot+scale_colour_brewer(palette = "Sequential") 
plot_revised2<-plot_revised+theme_dark()
plot_revised2
  
  
wine_by_quality <- mydata %>%
      group_by(quality) %>%
      summarise(mean_volatile.acidity = mean(volatile.acidity),
              median_volatile.acidity = median(volatile.acidity),
              n = n()) %>%
      ungroup() %>%
      arrange(quality)
    ggplot(aes(x = quality, y = median_volatile.acidity), 
           data = wine_by_quality) + 
      geom_bar(stat = 'identity') + 
      scale_x_continuous(breaks = seq(3, 8, 1))


```

The volatile acidity is lower for wines of higher quality.

```{r echo=FALSE,  message=FALSE, Include="FALSE", warning=FALSE,quality_vs_citirc_acid}
plot<-ggplot(aes(x = factor(quality), y = citric.acid), data = mydata) +
  geom_jitter(alpha=0.3)+
  geom_boxplot(alpha=0.5,color="blue",
               outlier.color=NA) +
  stat_summary(fun.y="mean", geom="point", color="red",
               shape=10, size=4 )
  facet_wrap(~quality, nrow=1) 
  
plot_revised<-plot+scale_colour_brewer(palette = "Sequential") 
plot_revised2<-plot_revised+theme_dark()
plot_revised2
  
 
wine_by_quality <- mydata %>%
      group_by(quality) %>%
      summarise(mean_citric.acid = mean(citric.acid),
              median_citric.acid = median(citric.acid),
              n = n()) %>%
      ungroup() %>%
      arrange(quality)
    ggplot(aes(x = quality, y = median_citric.acid), data = wine_by_quality) + 
      geom_bar(stat = 'identity') + 
      scale_x_continuous(breaks = seq(3, 8, 1))

cor.test(mydata$citric.acid, mydata$quality)

```

High quality wines have higher quantity of citric acid. Correlation is 0.22

```{r echo=FALSE, message=FALSE, include="FALSE", warning=FALSE,quality_vs_sulphates }
ggplot(aes(x = factor(quality), y =sulphates ), data = mydata) +
  geom_jitter(alpha=0.3)+
  geom_boxplot(alpha=0.5,color="blue",
               outlier.color=NA) +
  stat_summary(fun.y="mean", geom="point", color="red",
               shape=10, size=4 )
  facet_wrap(~quality, nrow=1) 
  
  
wine_by_quality <- mydata %>%
      group_by(quality) %>%
      summarise(mean_sulphates = mean(sulphates),
              median_sulphates = median(sulphates),
              n = n()) %>%
      ungroup() %>%
      arrange(quality)
    ggplot(aes(x = quality, y = median_sulphates), 
           data = wine_by_quality) + 
      geom_bar(stat = 'identity') + 
      scale_x_continuous(breaks = seq(3, 8, 1))
    
cor.test(mydata$sulphates, mydata$quality)
```

Again, the quality of wine is better if the sulphates content is higher. Correlation is 0.25


Let's try to see how the input variables may be related to each other. 
```{r echo=FALSE,  message=FALSE, warning=FALSE,free_vs_total}
ggplot(aes(x = free.sulfur.dioxide, y = total.sulfur.dioxide), 
       data = mydata) +
  geom_point(alpha=1/5, position="jitter") +
  scale_x_continuous(breaks = seq(3, 8, 1))
  
 cor.test(mydata$free.sulfur.dioxide, mydata$total.sulfur.dioxide)
  
```


 Looks slightly exponential. Correlation is 0.67 which still means they are correlated. This may be due to the fact that total sulfur dioxide contributes to free sulfur dioxide. For the same reason, I comined the two in a single varibale sulfur_dioxide 

# Multivariate Plots Section

```{r echo=FALSE,  message=FALSE, warning=FALSE,multivariate_plot_1}
ggplot(aes(x=quality, y=alcohol, color=citric.acid), 
       data=mydata)+
  geom_point(position="jitter")

```

nothing particulary stands out from this plot.

Perhaps, a multivariate plot makes more sense if plotted with a categorical variable. However, our dataset doesn't have any useful categorical varibale. But we can create one! 

```{r echo=FALSE, message=FALSE, warning=FALSE, bucketing_alcohol}
mydata$alcohol_level <- cut(mydata$alcohol, 
              breaks = c(quantile(mydata$alcohol, prob =   0),
              quantile(mydata$alcohol, prob = .25), 
              quantile(mydata$alcohol, prob = .50),
              quantile(mydata$alcohol, prob =  .75),
              quantile(mydata$alcohol, prob =1)),
labels = c("Low", "Medium", "High", "Very-High"))

```

I bucketed the alcohol variable to four groups-"Low", "Medium", "High", "Very-High") based on quantile methods

```{r echo=FALSE, message=FALSE, warning=FALSE, multivariate_plot_2}
plot<-ggplot(aes(x=quality, y=alcohol, color=alcohol_level), 
       data=mydata)+
  geom_point(position="jitter")

plot_revised<-plot+scale_colour_brewer(palette = "Sequential") 
plot_revised2<-plot_revised+theme_dark()
plot_revised2

```

We can see that towards the right side of the plot which has higher quality wine, the alcohol contnt is either high or very high.
But we already know this from our bivariate analysis.

Let's create another categorical variable by bucketing the ctric acid varibale
categorical varibale. But we can create one! 

```{r echo=FALSE,  message=FALSE, warning=FALSE,bucketing_citirc_acid}
mydata$citric.acid_level <- cut(mydata$citric.acid, 
              breaks = c(quantile(mydata$citric.acid, prob =   0),
              quantile(mydata$citric.acid, prob = .25), 
              quantile(mydata$citric.acid, prob = .50),
              quantile(mydata$citric.acid, prob =  .75),
              quantile(mydata$citric.acid, prob =1)),
labels = c("Low", "Medium", "High", "Very-High"))

```

```{r echo=FALSE,  message=FALSE, warning=FALSE,multivariate_plot_3}
plot<-ggplot(aes(x=quality, y=alcohol, color=citric.acid_level),
       data=mydata)+
  geom_point(position="jitter")

plot_revised<-plot+scale_colour_brewer(palette = "Sequential") 
plot_revised2<-plot_revised+theme_dark()
plot_revised2

```

There is nothing conclusive from this plot

Let's bucket sulphate 

```{r echo=FALSE, bucketing_sulphates}
mydata$sulphates_level <- cut(mydata$sulphates, 
              breaks = c(quantile(mydata$sulphates, prob =   0),
              quantile(mydata$sulphates, prob = .25), 
              quantile(mydata$sulphates, prob = .50),
              quantile(mydata$sulphates, prob =  .75),
              quantile(mydata$sulphates, prob =1)),
labels = c("Low", "Medium", "High", "Very-High"))

```

```{r echo=FALSE,  message=FALSE, warning=FALSE,multivariate_plot_4}
plot<-ggplot(aes(x=quality, y=alcohol, color=sulphates_level), 
       data=mydata)+
  geom_point(position="jitter")

plot_revised<-plot+scale_colour_brewer(palette = "Sequential") 
plot_revised2<-plot_revised+theme_dark()
plot_revised2

```

Though this plot is not very conclusive, we can see that the number of low and medium level dots significantly reduce in the high quality dots

Let's also check for volatile acidity:-

```{r echo=FALSE, message=FALSE, warning=FALSE, bucketing_volatile_acidity}
mydata$volatile.acidity_level <- cut(mydata$volatile.acidity, 
              breaks = c(quantile(mydata$volatile.acidity, prob =   0),
              quantile(mydata$volatile.acidity, prob = .25), 
              quantile(mydata$volatile.acidity, prob = .50),
              quantile(mydata$volatile.acidity, prob =  .75),
              quantile(mydata$volatile.acidity, prob =1)),
labels = c("Low", "Medium", "High", "Very-High"))

```

```{r echo=FALSE,  message=FALSE, warning=FALSE,multivariate_plot_5}
plot<-ggplot(aes(x=quality, y=alcohol, color=volatile.acidity_level), 
       data=mydata)+
  geom_point(position="jitter")

plot_revised<-plot+scale_colour_brewer(palette = "Sequential") 
plot_revised2<-plot_revised+theme_dark()
plot_revised2

```

Again from this plot we can only say that the volatile acidity is low for high quality wine.



------

# Final Plots and Summary

### Plot One: Understanding our dataset in terms of the main feature of interest, quality
```{r echo=FALSE,  message=FALSE, warning=FALSE,Plot_One}
first_plot<-ggplot(aes(x=quality), data=mydata)+ 
  geom_histogram(aes(y = ..density..), 
                 binwidth=1,colour="black", fill="orange" )+
  xlab("Quality( Scale 0-10)") + ylab("Density")+
  ggtitle("Distrbution of quality")
  scale_x_continuous(lim=c(3,8), breaks=seq(3,8,1))
  
first_plot + theme(
       title = element_text("Histogram showing frequency of wines based on quality score"),
       plot.title = element_text(face="bold", size=18), 
       axis.title.x = element_text(face="bold", size=16),
       axis.title.y = element_text(face="bold", size=14, angle=90),
       axis.text.x = element_text(face="bold", size=14),
       axis.text.y = element_text(face="bold", size=14)
)

```

### Description One
Plot one gives us a good insight of the dataset. It tells us that most of the wines are of average quality as the quality score lie between 5-7 on a scale of 0-10. It also tells us that the quality scores lie between 3-8. This information can be used to classify good quality wines and poor quality wines.

### Plot Two: A boxplot showing how alcohol content influences quality of wine
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Two}
second_plot<-ggplot(aes(x = factor(quality), y =alcohol ), data = mydata) +
  geom_jitter(alpha=0.3)+
  geom_boxplot(alpha=0.5,color="blue",
               outlier.color=NA) +
  stat_summary(fun.y="mean", geom="point", color="red",
               shape=10, size=4 )
 
  
second_plot+xlab("Quality( Scale 0-10)")+
  ylab("Alcohol content (% by Volume")+
  ggtitle("Alcohol content vs quality of wine")+ 
  theme(
       title = element_text("Histogram showing frequency of wines based on quality score"),
       plot.title = element_text(face="bold", size=18), 
       axis.title.x = element_text(face="bold", size=16),
       axis.title.y = element_text(face="bold",
                                   size=14, angle=90),
       axis.text.x = element_text(face="bold", size=14),
       axis.text.y = element_text(face="bold", size=14)
)+
  scale_colour_brewer(palette = "Sequential") + 
  theme_dark()
  
```

### Description Two
This plot shows that that alcohol content and quality of wine are positively correlated. As the alcohol content increases, which is the median of alcohol content of wines grouped by quality score, the quality of wine also increases. This is confirmed by the Pearson correlation between the two variables which is 0.45. In fact, this is the highest correlation of the quality with any other input variable. Therefore, alcohol content seems to be the most infuential psoitive character.


### Plot Three: Check if a thrid variable supports the correlation between alcohol and qaulity and wine
```{r echo=FALSE,  message=FALSE, warning=FALSE,Plot_Three}
third_plot<-ggplot(aes(x=quality, y=alcohol, 
                       color=volatile.acidity_level),
                   data=mydata)+
  geom_point(position="jitter")
 
                        
                     
  
third_plot+xlab("Quality( Scale 0-10)")+
  ylab("Alcohol content (% by Volume)")+
  ggtitle("Effect of Volatile Acidity")+ 
  theme(
       title = element_text("Alcohol content vs quality of wine  with volatile acidity as a categorical variable"),
       plot.title = element_text(face="bold", size=18), 
       axis.title.x = element_text(face="bold", size=16),
       axis.title.y = element_text(face="bold", size=14, angle=90),
       axis.text.x = element_text(face="bold", size=14),
       axis.text.y = element_text(face="bold", size=14)
)+ scale_colour_brewer(palette = "Sequential", guide = guide_legend(title = 'Volatile acidity (g/dm3)')) + 
  theme_dark()
 


```

### Description Three
This plot is an attempt to understand if the effect of a thrid variable can be included to support the correlation between two varibales. A categorical variable was created by bucketing the variable volatile acidity using quantile methods into four categories "Low", "Medium", "High", "Very-High". The reason to pick the volatile acidity is because it is negatively correlated to the quality and hence it would be good to have it on the same plot as a positive factor(alcohol).
However, the plot we see here is not very good to draw any strong conclusions about the trend. This tells us that the relation is nonlinear. We can only get a basic idea that the quality of wine improves when the volatile acidity goes down.
------

# Reflection
Overall, this was an interesting data set to work with. We were ineterested in exploring how different input variables influence the output variable-quality. The first important decision I made that helped with the analyses was to find the Pearson correlation of quality with all other variables. This gave an idea of the factors that are contributing most to the quality of wine. These are, alcohol content, volatile acidity, citric acod content and the sulphates content. 
The main problem I encountered was I wasn't able to find a strong linear correlation between the varibales.
This can be overcome by using support vector machine model as it the best for small data sets with non linear trends.
The prodictive model would greatly enhance the quality of data analysis and I would like to work on it in the future.
